---
title: "Scalable Generalized Mixed Models in PheWAS using SAIGEgds"
author: "Xiuwen Zheng"
date: "Sep, 2026"
header-includes:
    - \usepackage{array}
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
        number_sections: true
    BiocStyle::pdf_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{SAIGEgds Tutorial (single variant tests)}
    %\VignetteDepends{SAIGEgds}
    %\VignetteKeywords{GWAS, PheWAS, mixed models}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

# Introduction

A phenome-wide association study (PheWAS) is a powerful tool for the discovery and replication of genetic associations across many phenotypes. Large biobanks provide deep genomic and phenotypic data, but testing thousands of traits against whole-genome variants remains computationally demanding. SAIGE (scalable and accurate implementation of generalized mixed models [2]) addresses case-control imbalance and sample relatedness in association studies.

SAIGEgds is a high-performance implementation of the SAIGE framework using Genomic Data Structure (GDS) files [1,3]. Its optimized C++ code takes advantage of sparse genotype dosages and supports integer genotypes and numeric imputed dosages. SAIGEgds supports binary, quantitative and time-to-event outcomes, as well as single-variant and set-based association tests. This vignette focuses on the binary-trait, single-variant workflow; time-to-event outcomes are covered in the companion survival vignette.

Benchmarks using UK Biobank White British genotype data ($N=430\mathrm{K}$) with coronary heart disease and simulated case-control outcomes showed that SAIGEgds was 5 to 6 times faster than the SAIGE R package for null-model fitting and p-value calculation. Together with high-performance computing (HPC) clusters or cloud resources, it provides an efficient pipeline for biobank-scale PheWAS.

# Installation

* Bioconductor repository

```R
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("SAIGEgds")
```
The `BiocManager::install()` approach may build the package from source. In that case, `make` and a suitable compiler toolchain must be installed; see the [R FAQ](https://cran.r-project.org/faqs.html) for platform-specific instructions.

# Workflow

```{r message=FALSE}
library(SeqArray)
library(SAIGEgds)

# the genotype file in SeqArray GDS format (1000 Genomes Phase1, chromosome 22)
(geno_fn <- seqExampleFileName("KG_Phase1"))
```

## Preparing SNP data for the genetic relationship matrix

```{r}
# open a SeqArray file in the package (1000 Genomes Phase1, chromosome 22)
gds <- seqOpen(geno_fn)
```

The LD pruning is provided by [snpgdsLDpruning()](https://rdrr.io/bioc/SNPRelate/man/snpgdsLDpruning.html) in the package [SNPRelate](https://bioconductor.org/packages/SNPRelate):
```{r}
library(SNPRelate)

set.seed(1000)
snpset <- snpgdsLDpruning(gds)
str(snpset)

snpset.id <- unlist(snpset, use.names=FALSE)  # get the variant IDs of a LD-pruned set
head(snpset.id)
```

Create a genotype file for the genetic relationship matrix (GRM) using the LD-pruned SNP set:
```{r}
grm_fn <- "grm_geno.gds"
seqSetFilter(gds, variant.id=snpset.id)

# export to a GDS genotype file without annotation data
seqExport(gds, grm_fn, info.var=character(), fmt.var=character(), samp.var=character())
```

If genotypes are split by chromosomes, `seqMerge()` in the SeqArray package can be used to combine the LD-pruned SNP sets.

```{r}
# close the file
seqClose(gds)
```

## Fitting the null model

A simulated phenotype data set is used to demonstrate model fitting. `sample.col` identifies the phenotype column that is matched to sample IDs in the GDS file; samples without complete model variables or without matching genotypes are excluded.
```{r}
set.seed(1000)
sampid <- seqGetData(grm_fn, "sample.id")  # sample IDs in the genotype file

pheno <- data.frame(
    sample.id = sampid,
    y  = sample(c(0, 1), length(sampid), replace=TRUE, prob=c(0.95, 0.05)),
    x1 = rnorm(length(sampid)),
    x2 = rnorm(length(sampid)),
    stringsAsFactors = FALSE)
head(pheno)

grm_fn
```

Null-model fitting is the computationally intensive step. The call is shown below but is not run while building this vignette:

```{r eval=FALSE}
# null model fitting using GRM from grm_fn
glmm <- seqFitNullGLMM_SPA(y ~ x1 + x2, pheno, grm_fn, trait.type="binary", sample.col="sample.id")
```

For a fast and reproducible vignette build, the package includes a model precomputed with the same workflow:

```{r echo=FALSE}
glmm <- readRDS(system.file("extdata", "v_glmm.rds", package="SAIGEgds"))
```

Check the sample count, convergence status and estimated variance components before running association tests:

```{r}
length(glmm$sample.id)
glmm$converged
glmm$tau
```

Note what the components of a binary null model mean:

| Component | Meaning for `trait.type="binary"` |
|:--|:--|
| `coefficients` | fixed effects on the **log-odds** scale, including the intercept |
| `tau` | `Sigma_E` is fixed at 1, and `Sigma_G` is the variance of the genetic random effect on the log-odds scale |
| `fitted.values` | $\mu_i = \Pr(y_i=1)$, the fitted probability of the outcome |
| `residuals` | $y_i - \mu_i$, the raw residuals |
| `var.ratio` | variance ratio(s) used to approximate the score-statistic variance in association tests |

: Components of a binary null model.

## P-value calculations

```{r}
# genetic variants stored in the file geno_fn
geno_fn
# use parallel > 1 in production to distribute variants across processes
assoc <- seqAssocGLMM_SPA(geno_fn, glmm, mac=10, parallel=1)

head(assoc)

# basic result checks
c(n.variant=nrow(assoc), n.finite.p=sum(is.finite(assoc$pval)))

# filtering based on p-value
assoc[assoc$pval < 5e-4, ]
```

The main result columns report the variant location and allele frequency, effect estimate (`beta`), standard error (`SE`) and p-value (`pval`). The `method` and `converged` columns indicate which p-value approximation was used and whether its numerical calculation converged.

The output can also be saved directly to an R object file or a GDS file:
```{r}
# save to 'assoc.gds'
seqAssocGLMM_SPA(geno_fn, glmm, mac=10, parallel=1, res.savefn="assoc.gds")
```

Open the output GDS file using the functions in the [gdsfmt](https://www.bioconductor.org/packages/gdsfmt) package:
```{r}
# open the GDS file
(f <- openfn.gds("assoc.gds"))
# get p-values
pval <- read.gdsn(index.gdsn(f, "pval"))
summary(pval)
closefn.gds(f)
```

Load association results using the function `seqSAIGE_LoadPval()` in SAIGEgds:
```{r}
res <- seqSAIGE_LoadPval("assoc.gds")
head(res)
```

## Manhattan and QQ plots for p-values

```{r fig.width=6, fig.height=3, fig.align='center'}
library(ggmanh)

g <- manhattan_plot(assoc, pval.colname="pval", chr.colname="chr",
    pos.colname="pos", x.label="Chromosome 22")
g
```

```{r fig.width=3, fig.height=3, fig.align='center'}
# QQ plot
qqunif(assoc$pval)
```

# GPU Acceleration (Optional)

SAIGEgds supports optional OpenCL-based GPU acceleration for the GRM cross-product computation during null-model fitting. This can significantly speed up `seqFitNullGLMM_SPA()` for large sample sizes with dense genotypes.

**Requirements:**

* An OpenCL-compatible GPU (NVIDIA, AMD, Intel, or Apple)
* OpenCL runtime library installed on the system (no compile-time dependency)

To enable GPU acceleration, set `use.gpu=TRUE`:
```R
glmm <- seqFitNullGLMM_SPA(y ~ x1 + x2, pheno, grm_fn,
    trait.type="binary", sample.col="sample.id", use.gpu=TRUE)
```

If no compatible GPU or OpenCL runtime is found, the function automatically falls back to CPU computation. When GPU initialization succeeds, `use.gpu=TRUE` automatically selects dense 2-bit packed genotype storage; users do not need to set `geno.sparse=FALSE` separately. The default sparse genotype path remains CPU-only and may be preferable when genotypes are sparse.

**GPU Memory Usage:**

The GPU memory required depends on the operation and dataset size. SAIGEgds checks that the total allocation does not exceed 80% of device global memory, and falls back to CPU if it does.

- **Cross-product (null-model fitting):** $\lceil N/4 \rceil M$ bytes (packed genotypes) + $16M$ bytes (standardized genotype lookup) + $8N + 4M$ bytes (input, output and dot-product vectors), where $N$ is the number of samples and $M$ is the number of variants. For example, $N=400\mathrm{K}$ and $M=100\mathrm{K}$ requires approximately 10 GB.
- **Dense GRM GEMV:** $4N^2$ bytes (GRM matrix in single precision) + $8N$ bytes (input/output vectors). For example, $N=5\mathrm{K}$ requires approximately 100 MB.
- **GRM construction (sparse/dense):** $\lceil M/4 \rceil N$ bytes (packed genotypes) + $16M$ bytes (standardized genotype lookup) + $4B^2$ bytes (an implementation-defined block output buffer of width $B$). For $N=400\mathrm{K}$ and $M=2\mathrm{K}$ (the default `nsnp.sub.random` subset), the packed genotypes require approximately 200 MB.

GPU compute buffers use single-precision (32-bit) floating point. Results are converted back to double precision on the host.


# Session Information

```{r echo=FALSE}
unlink(c("grm_geno.gds", "assoc.gds"), force=TRUE)
```

```{r}
sessionInfo()
```


# References

1. Zheng X, Davis JW. SAIGEgds -- an efficient statistical tool for large-scale PheWAS with mixed models. *Bioinformatics* (2021). Mar;37(5):728-730. [DOI: 10.1093/bioinformatics/btaa731](https://doi.org/10.1093/bioinformatics/btaa731).
2. Zhou W, Nielsen JB, Fritsche LG, Dey R, Gabrielsen ME, Wolford BN, LeFaive J, VandeHaar P, Gagliano SA, Gifford A, Bastarache LA, Wei WQ, Denny JC, Lin M, Hveem K, Kang HM, Abecasis GR, Willer CJ, Lee S. Efficiently controlling for case-control imbalance and sample relatedness in large-scale genetic association studies. *Nat Genet* (2018). Sep;50(9):1335-1341. [DOI: 10.1038/s41588-018-0184-y](https://doi.org/10.1038/s41588-018-0184-y)
3. Zheng X, Gogarten S, Lawrence M, Stilp A, Conomos M, Weir BS, Laurie C, Levine D. SeqArray -- A storage-efficient high-performance data format for WGS variant calls. *Bioinformatics* (2017). [DOI: 10.1093/bioinformatics/btx145](https://doi.org/10.1093/bioinformatics/btx145).


# See also

Time-to-event outcomes are covered in *Time-to-event (survival) analysis in SAIGEgds* (`vignette("SAIGEgds_surv", package="SAIGEgds")`). Rare-variant set-based tests using full and sparse GRMs are covered in *Set-based Rare-Variant Association Analysis in SAIGEgds* (`vignette("SAIGEgds_set-aggr", package="SAIGEgds")`).

[SeqArray](https://www.bioconductor.org/packages/SeqArray): Data Management of Large-scale Whole-genome Sequence Variant Calls

[SNPRelate](https://www.bioconductor.org/packages/SNPRelate): Parallel Computing Toolset for Relatedness and Principal Component Analysis of SNP Data
